---
title: "TidyTuesday 2018 - Week 23 - Fast Food Calories"
author: "Roberto Preste"
date: "2018-09-04"
output: html_document
editor_options: 
  chunk_output_type: inline
---

This is my work for week 23 (2018) of the [#TidyTuesday](https://thomasmock.netlify.com/post/tidytuesday-a-weekly-social-data-project-in-r/) project.

This week’s dataset was about nutritional information from several fast food franchises, with data regarding entrees food in particular. Since I am a bit busy at work after the summer holidays, I haven’t had much time to deeply explore this dataset. A more complete analysis can be found in the in the [original article](https://www.franchiseopportunities.com/blog/general-franchise-information/fast-food-calorie-comparison-charts), or by visiting [fastfoodnutrition.org](https://fastfoodnutrition.org/), which contains more extensive data on this topic.

All code and data can be found in my dedicated GitHub repository [MyTidyTuesday](https://github.com/robertopreste/MyTidyTuesday).

___

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 120)
```

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(magrittr)
library(rvest)
library(skimr)
```

___ 

Original data come from [fastfoodnutrition.org](https://fastfoodnutrition.org), and the original article about fast food calories comparison is available on [Franchise Opportunities](https://www.franchiseopportunities.com/blog/general-franchise-information/fast-food-calorie-comparison-charts).  

Firstly, we need to scrape the data from the data source (function kindly provided by [thomas_mock](https://twitter.com/thomas_mock)), then we can combine all the data and save them.  

```{r, eval=FALSE}
food_scrape <- function(restaurant, tbl_sel){
    url <- glue::glue("https://fastfoodnutrition.org/{restaurant}/chart")
    url %>% 
        read_html() %>% 
        html_table() %>% 
        .[tbl_sel] %>% 
        bind_rows() %>% 
        select(-X16) %>%
        set_names(nm = c("item", "calories", "cal_fat", "total_fat", "sat_fat", "trans_fat",
                         "cholesterol", "sodium", "total_carb", "fiber", "sugar", "protein",
                         "vit_a", "vit_c", "calcium")) %>% 
        mutate(restaurant = str_replace(restaurant, "-", " "),
               restaurant = str_to_title(restaurant)) # save restaurant name
} 
```

```{r, eval=FALSE}
mcd_df <- food_scrape("mcdonalds", c(1,2,3,9,19))
cfa_df <- food_scrape("chick-fil-a", c(1,2,8,13))
sonic_df <- food_scrape("sonic", c(1,2,17,18,20))
arbys_df <- food_scrape("arbys", c(1:4,8))
bk_df <- food_scrape("burger-king", c(1:3,7,11:12))
dq_df <- food_scrape("dairy-queen", c(5, 7, 12, 25, 27))
sub_df <- food_scrape("subway", c(1,2,3,4,5,6,7,8,9))
taco_df <- food_scrape("taco-bell", c(1,2,3,4,5,15,18,19,20,22,23,24))

final_df <- bind_rows(mcd_df, cfa_df, sonic_df, arbys_df, bk_df, dq_df, sub_df, taco_df) %>% 
    select(restaurant, everything()) %>% 
    mutate(salad = case_when(str_detect(item, "salad") ~ "Salad",
                             TRUE ~ "Other"))

final_df %>% write_csv("data/fastfood_calories.csv")
```

The final dataframe is available in `data/fastfood_calories.csv`.  
It contains nutritional information about entrees (main courses) from the specified fast foot franchises.  

```{r, results="hide"}
df <- read_csv("data/fastfood_calories.csv")
```

___

## Data exploration  

Let's have a quick view at the data.

```{r}
head(df)
```

```{r}
skim(df)
```


We can drop the `salad` feature, because it has the same value (`Other`) for all the observations in the data, so it's uninformative.  

```{r}
df %<>% select(-salad)
```

___

## Visualization  

I chose to focus just on a couple of nutrients from these data, namely some of the most *feared* ones: fat and sugar.  

### Fat content per franchise  

Let's visualize the general fat content of entrees per franchise, distinguished in total grams of fat, saturated fat and trans fat.  

```{r}
fat_content <- df %>% 
    group_by(restaurant) %>% 
    summarise(Fat = mean(total_fat), 
              Saturated = mean(sat_fat), 
              Trans = mean(trans_fat)) %>% 
    gather(measure, value, Fat:Trans)
```

```{r, dpi=200}
fat_content %>% 
    ggplot(aes(x = reorder(restaurant, value))) + 
    geom_col(aes(y = value, fill = measure), position = "dodge") +
    coord_flip() + 
    labs(x = "Franchise", y = "Fat (g)", title = "Mean fat content per franchise", fill = "", subtitle = "Chick-Fil-A seems to have low-fat food.") 
```

Seems like, in general, Chick-Fil-A might offer food with a lower fat content, with Subway and Taco Bell immediately following. Choosing one of these three franchises can be two to three times an healthier options compared to other fast foods.  


### Sugar content per franchise  

Another bad player for our health is sugar, so let's check how well our fast foods behave.  

```{r}
sugar_content <- df %>% 
    group_by(restaurant) %>% 
    summarise(Sugar = mean(sugar))
```

```{r, dpi=200}
sugar_content %>% 
    ggplot(aes(x = reorder(restaurant, Sugar), y = Sugar, fill = restaurant)) + 
    geom_col() + 
    coord_flip() + 
    labs(x = "Franchise", y = "Sugar (g)", title = "Mean sugar content per franchise", subtitle = "Taco Bell might have low-sugar food.") + 
    guides(fill = FALSE)
```

A histogram can give a better overview of data and show possible outliers, represented by food with extremely high content of sugar.  

```{r, dpi=200}
df %>% 
    ggplot(aes(x = sugar, fill = restaurant)) + 
    geom_histogram(bins = 60) + 
    facet_wrap(~ restaurant, nrow = 4, ncol = 2) + 
    labs(x = "Sugar (g)", title = "Sugar content distribution per franchise") +
    guides(fill = FALSE)
```

Let's zoom in a bit and remove the above-mentioned outliers.  

```{r, dpi=200}
df %>% 
    filter(sugar <= 30) %>% 
    ggplot(aes(x = sugar, fill = restaurant)) + 
    geom_histogram(bins = 30) + 
    facet_wrap(~ restaurant, nrow = 4, ncol = 2) + 
    labs(x = "Sugar (g)", title = "Sugar content distribution per franchise [0-30 g]") +
    guides(fill = FALSE)
```

An interesting feature can be noted here: Taco Bell offers food with the lowest sugar content, not exceeding 9 grams, while other franchises span a wide range of sugar content. Nonetheless, seems like Taco Bell, together with Subway, do not offer sugar-free options as other fast foods do.  

___

### Disclaimer  

These are just some basic insights, created for simple data exploration and visualization purposes. I'm not a nutritionist, and these are not comprehensive data from which any useful and trustworthy information can be extracted.  
No conclusions should be drawn from what is reported here.  

___

```{r}
sessionInfo()
```