---
title: "TidyTuesday 2018 - Week 19 - Airline Safety"
author: "Roberto Preste"
date: "2018-08-07"
output: html_document
---

This is an extract from my work for week 19 (2018) of the [#TidyTuesday](https://thomasmock.netlify.com/post/tidytuesday-a-weekly-social-data-project-in-r/) project.

This week’s dataset was focused on airlines’ accidents, particularly on differences between the 1985-1999 and 2000-2014 years. I decided to highlight how much each airline company reduced (or not) the amount of incidents, fatalities and fatal accidents in recent years. More information on these categories of accidents can be found in the [original article](https://fivethirtyeight.com/features/should-travelers-avoid-flying-airlines-that-have-had-crashes-in-the-past/).

All code and data can be found in my dedicated GitHub repository [MyTidyTuesday](https://github.com/robertopreste/MyTidyTuesday).

___

```{r}
library(tidyverse)
library(fivethirtyeight)
data("airline_safety")
```

___

These are the starting data (the original article can be found [here](https://fivethirtyeight.com/features/should-travelers-avoid-flying-airlines-that-have-had-crashes-in-the-past/)):  

```{r}
head(airline_safety)
```

___

## Tidying the data  

Let's calculate the difference of accidents in 2000-2014 vs 1985-1999; lower values mean a reduced number of accidents in recent years. After that, we'll `gather` these values.  

```{r}
airline_diff <- airline_safety %>% 
    mutate(fatal_accidents = fatal_accidents_00_14 - fatal_accidents_85_99, 
           fatalities = fatalities_00_14 - fatalities_85_99, 
           incidents = incidents_00_14 - incidents_85_99) %>% 
    gather(key = "event", value = "occurrences", fatal_accidents, fatalities, incidents) %>% 
    select(everything(), -c(fatal_accidents_85_99, fatal_accidents_00_14, fatalities_85_99, fatalities_00_14, incidents_85_99, incidents_00_14))
```

The tidy dataset looks like this:  

```{r}
head(airline_diff)
```

___

## Visualizations  


```{r dpi=200}
airline_diff %>% 
    filter(event == "fatalities", occurrences != 0) %>% 
    ggplot(aes(x = reorder(airline, occurrences), y = occurrences, fill = occurrences)) + 
    geom_col() + 
    coord_flip() + 
    scale_fill_gradientn(colors = c("darkgreen", "aquamarine3", "seagreen3", "yellow", "orange", "darkred")) +
    labs(x = "Airline", y = "Fatalities", fill = "", title = "Difference in number of fatalities", subtitle = "Years 1985-1999 vs 2000-2014")
```

```{r dpi=200}
airline_diff %>% 
    filter(event == "fatal_accidents", occurrences != 0) %>% 
    ggplot(aes(x = reorder(airline, occurrences), y = occurrences, fill = occurrences)) + 
    geom_col() + 
    coord_flip() +
    scale_fill_gradientn(colors = c("darkgreen", "aquamarine3", "seagreen3", "orange", "darkred"), values = c(0, 0.6, 0.7, 0.8, 1)) + 
    labs(x = "Airline", y = "Fatal Accidents", fill = "", title = "Difference in number of fatal accidents", subtitle = "Years 1985-1999 vs 2000-2014")
```

```{r dpi=200}
airline_diff %>% 
    filter(event == "incidents", occurrences != 0) %>% 
    ggplot(aes(x = reorder(airline, occurrences), y = occurrences, fill = occurrences)) + 
    geom_col() + 
    coord_flip() +
    scale_fill_gradientn(colors = c("darkgreen", "aquamarine3", "seagreen3", "orange", "darkred"), values = c(0, 0.7, 0.8, 0.9, 1)) + 
    labs(x = "Airline", y = "Incidents", fill = "", title = "Difference in number of incidents", subtitle = "Years 1985-1999 vs 2000-2014")
```

___  

```{r}
sessionInfo()
```

