---
title: "TidyTuesday Week 21 - California Fires"
author: "Roberto Preste"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(funModeling)
library(gridExtra)
```

___ 

This is the original data provided by [BuzzFeed](https://github.com/BuzzFeedNews/2018-07-wildfire-trends) from their article ["How A Booming Population And Climate Change Made California’s Wildfires Worse Than Ever"](https://www.buzzfeednews.com/article/peteraldhous/california-wildfires-people-climate).  

```{r, results='hide'}
df = read_csv("data/calfire_frap.csv", skip = 1, 
              col_names = c("id", "objectid", "year", "state", "agency", "unit_id", "fire_name", "inc_num", 
                            "alarm_date", "cont_date", "cause", "comments", "report_ac", "gis_acres", "c_method", 
                            "objective", "fire_num", "shape_length", "shape_area", "fire_cause", "plot_date"))
```

```{r}
head(df)
```

___  

## Data exploration  

Let's have a quick view at the structure and content of this dataset.  

```{r}
df %>% df_status()
```


Let's first check the number of human and natural fires throughout the years.  

```{r}
fire_per_years <- df %>% 
    group_by(year, fire_cause) %>% 
    summarise(fires = n(), 
              tot_area_burnt = sum(shape_area), 
              mean_area_burnt = mean(shape_area))
```

We will plot the same data using a simple line plot and a smooth line plot, to avoid overplotting and visualize better the trends.  

```{r, fig.height=4}
line_1 <- fire_type_years %>% 
    ggplot(aes(x = year, y = fires, colour = fire_cause)) + 
    geom_line()
smooth_1 <- fire_type_years %>% 
    ggplot(aes(x = year, y = fires, colour = fire_cause)) + 
    geom_smooth()
grid.arrange(line_1, smooth_1, nrow = 2)
```

```{r, fig.width=4}
fire_type_years %>% 
    ggplot(aes(x = year, y = fires)) + 
    geom_col(aes(fill = fire_cause), position = "dodge") +
    geom_smooth(aes(color = fire_cause)) + 
    facet_grid(~ fire_cause) + 
    guides(fill = FALSE, color = FALSE)
```


We can clearly identify that in recent years, human-caused fires have definitely outnumbered the natural ones!  

Let's see if this increase also regards the burnt area of a fire.  

```{r}
fire_area_years <- df %>% 
    group_by(year, fire_cause) %>% 
    summarise(tot_area_burnt = sum(shape_area), 
              mean_area_burnt = mean(shape_area))
```

```{r, fig.height=4}
line_2 <- fire_area_years %>% 
    ggplot(aes(x = year, y = tot_area_burnt, color = fire_cause)) + 
    geom_line()
smooth_2 <- fire_area_years %>% 
    ggplot(aes(x = year, y = tot_area_burnt, color = fire_cause)) + 
    geom_smooth()
grid.arrange(line_2, smooth_2, nrow = 2)
```

Unfortunately, also the total area burnt each year increased over the years, although with no visible difference between natural- and human-caused fires.  
We may interested in the **mean** area burnt by each fire, though.  

```{r, fig.height=4}
line_3 <- fire_area_years %>% 
    ggplot(aes(x = year, y = mean_area_burnt, colour = fire_cause)) + 
    geom_line()
smooth_3 <- fire_area_years %>% 
    ggplot(aes(x = year, y = mean_area_burnt, colour = fire_cause)) + 
    geom_smooth()
grid.arrange(line_3, smooth_3, nrow = 2)
```

From the first line plot, we can see that the mean area burnt by each fire was roughly the same over the years; in this case the smooth plot is not very useful, because it's biased by the two peaks visible in the first graph. They are due to the 2002 [Biscuit Fire](https://en.wikipedia.org/wiki/Biscuit_Fire) and the 2012 [Rush Fire](https://en.wikipedia.org/wiki/Rush_Fire), which destroyed an impressive number of acres during their life.  
Notably, the Biscuit fire caused more destruction alone than all the other fires (natural and human) in 2002!  

```{r}
biscuit_area <- df %>% 
    filter(fire_name == "BISCUIT", year == 2002) %>% 
    select(shape_area)
others_area <- df %>% 
    filter(fire_name != "BISCUIT", fire_name != "RUSH", year == 2002) %>% 
    summarise(tot = sum(shape_area))

area_comp <- tibble(fire_name = c("Biscuit", "Others"), 
                    area_burnt = c(biscuit_area[[1]], others_area[[1]]))
```

```{r}
area_comp %>% 
    ggplot(aes(x = fire_name, y = area_burnt, fill = fire_name)) + 
    geom_col()
```



```{r}
fire_area_years %>% filter(year == 2012)
```

```{r}
df %>% filter(year == 2012) %>% arrange(desc(shape_area))
```


```{r}
df %>% select(alarm_date) %>% is.na()
```

